---
title: "Setting DESeq contrasts (using the model matrix)"
output: 
  md_document:
    variant: markdown_github
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(10)
```

```{r, message=FALSE, echo=FALSE}
library(DESeq2)
```

In this tutorial we show how to set treatment contrasts in `<DESeq2>` using the design or model matrix. 
This is a general and flexible way to define contrasts, and is often useful for more complex contrasts or when the design of the experiment is imbalanced (e.g. different number of replicates in each group). 
Although we focus on `<DESeq2>`, the approach can also be used with the other popular package `<edgeR>`.

Each section below covers a particular experimental design, from simpler to more complex ones. 
The first chunk of code in each section is to simulate data, which has no particular meaning and is only done in order to have a DESeqDataSet object with the right kind of variables for each example. 
In practice, users can ignore this step as they should have created a DESeqDataSet object from their own data following the [instructions in the vignette](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#the-deseqdataset). 

There is a set of [accompanying slides](https://docs.google.com/presentation/d/1B9zW1_F-kBqQEu4xqxIJrudYP5DecytYMRR6bY4H6aM/edit?usp=sharing) that illustrate each of the sections below. 


# One factor, two levels (slide 5)

```{r}
# simulate data
dds <- makeExampleDESeqDataSet(n = 1000, m = 6, betaSD = 2)
dds$condition <- factor(rep(c("shade", "sun"), each = 3))
```

First we can look at our sample information:

```{r}
colData(dds)
```

Our factor of interest is `condition` and so we define our design and run the DESeq model fitting routine:

```{r, message=FALSE}
design(dds) <- ~ 1 + condition # or just `~ condition`
dds <- DESeq(dds) # equivalent to edgeR::glmFit()
```

Then check what coefficients DESeq estimated:

```{r}
resultsNames(dds)
```

We can see that we have a coefficient for our _intercept_ and coefficient for the effect of "sun" (i.e. differences between sun versus shade). 

Using the more standard syntax, we can obtain the results for the effect of sun as such:

```{r}
res1 <- results(dds, contrast = list("condition_sun_vs_shade"))
res1
```

The above is a simple way to obtain the results of interest. 
But it is worth understanding how DESeq is getting to these results by looking at the model's matrix. 
DESeq defines the model matrix using base R functionality:

```{r}
model.matrix(design(dds), colData(dds))
```

We can see that R coded "condition" as a dummy variable, with an intercept (common to all samples) and a "conditionsun" variable, which adds the effect of sun to samples 4-6. 

We can actually set our contrasts in `DESeq2::results()` using a numeric vector. 
The way it works is to define a vector of "weights" for the coefficient(s) we want to test for.
In this case, we have `(Intercept)` and `conditionsun` as our coefficients (see model matrix above), and we want to test for the effect of sun, so our contrast vector would be `c(0, 1)`. 
In other words, we don't care about the value of `(Intercept)` (so it has a weight of 0), and we're only interested in the effect of sun (so we give it a weight of 1). 

In this case the design is very simple, so we could define our contrast vector "manually". But for complex designs this can get more difficult to do, so it's worth mentioning the general way in which we can define this. 
For any contrast of interest, we can follow three steps:

- Get the model matrix
- Subset the matrix for each group of interest and calculate its column means - this results in a vector of coefficients for each group
- Subtract the group vectors from each other according to the comparison we're interested in

Let's see this example in action:

```{r}
# get the model matrix
mod_mat <- model.matrix(design(dds), colData(dds))
mod_mat
```

```{r}
# calculate the vector of coefficient weights in the sun
sun <- colMeans(mod_mat[dds$condition == "sun", ])
sun
```

```{r}
# calculate the vector of coefficient weights in the shade
shade <- colMeans(mod_mat[dds$condition == "shade", ])
shade
```

```{r}
# The contrast we are interested in is the difference between sun and shade
sun - shade
```

That last step is where we define our contrast vector, and we can give this directly to the `results` function:

```{r}
# get the results for this contrast
res2 <- results(dds, contrast = sun - shade)
```

This gives us exactly the same results as before, which we can check for example by plotting the log-fold-changes between the first and second approach:

```{r, eval=FALSE}
plot(res1$log2FoldChange, res2$log2FoldChange)
```


## Extra: recoding the design (slide 12)

Often, we can use different model matrices that essentially correspond to the same design. 
For example, we could recode our design above by removing the intercept:

```{r, message=FALSE}
design(dds) <- ~ 0 + condition
dds <- DESeq(dds)
resultsNames(dds)
```

In this case we get a coefficient corresponding to the average expression in shade and the average expression in the sun (rather than the _difference_ between sun and shade).

If we use the same contrast trick as before (using the model matrix), we can see the result is the same:

```{r}
# get the model matrix
mod_mat <- model.matrix(design(dds), colData(dds))
mod_mat

# calculate weights for coefficients in each condition
sun <- colMeans(mod_mat[which(dds$condition == "sun"), ])
shade <- colMeans(mod_mat[which(dds$condition == "shade"), ])

# get the results for our contrast
res3 <- results(dds, contrast = sun - shade)
```

Again, the results are essentially the same:

```{r, eval=FALSE}
plot(res1$log2FoldChange, res3$log2FoldChange)
```

In theory there's no difference between these two ways of defining our design. 
The design with an intercept is more common, but for the purposes of understanding what's going on, it's sometimes easier to look at models without intercept. 


# One factor, three levels (slide 6)

```{r}
# simulate data
dds <- makeExampleDESeqDataSet(n = 1000, m = 9, betaSD = 2)
dds$condition <- NULL
dds$colour <- factor(rep(c("pink", "yellow", "white"), each = 3))
dds$colour <- relevel(dds$colour, "white")
```

First we can look at our sample information:

```{r}
colData(dds)
```

As in the previous example, we only have one factor of interest, `condition`, and so we define our design and run the DESeq as before:

```{r, message=FALSE}
design(dds) <- ~ 1 + colour
dds <- DESeq(dds)

# check the coefficients estimated by DEseq
resultsNames(dds)
```

We see that now we have 3 coefficients: 

- "Intercept" corresponds to white colour (our reference level)
- "colour_pink_vs_white" corresponds to the difference between the reference level and pink
- "colour_yellow_vs_white" corresponds to the difference between the reference level and yellow

We could obtain the difference between white and any of the two colours easily:

```{r}
res1_pink_white <- results(dds, contrast = list("colour_pink_vs_white"))
res1_yellow_white <- results(dds, contrast = list("colour_yellow_vs_white"))
```

For comparing pink vs yellow, however, we need to compare two coefficients with each other to check whether they are themselves different (check the slide to see the illustration). This is how the standard DESeq syntax would be:

```{r}
res1_pink_yellow <- results(dds, contrast = list("colour_pink_vs_white", 
                                                 "colour_yellow_vs_white"))
```

However, following our three steps detailed in the first section, we can define our comparisons from the design matrix:

```{r}
# define the model matrix
mod_mat <- model.matrix(design(dds), colData(dds))
mod_mat
```

```{r}
# calculate coefficient vectors for each group
pink <- colMeans(mod_mat[dds$colour == "pink", ])
white <- colMeans(mod_mat[dds$colour == "white", ])
yellow <- colMeans(mod_mat[dds$colour == "yellow", ])
```

And we can now define any contrasts we want:

```{r, fig.keep = "none"}
# obtain results for each pairwise contrast
res2_pink_white <- results(dds, contrast = pink - white)
res2_pink_yellow <- results(dds, contrast = pink - yellow)
res2_yellow_white <- results(dds, contrast = yellow - white)

# plot the results from the two approaches to check that they are identical
plot(res1_pink_white$log2FoldChange, res2_pink_white$log2FoldChange)
plot(res1_pink_yellow$log2FoldChange, res2_pink_yellow$log2FoldChange)
plot(res1_yellow_white$log2FoldChange, res2_yellow_white$log2FoldChange)
```

With this approach, we could even define a more unusual contrast, for example to find genes that differ between pigmented and non-pigmented samples:

```{r}
# define vector of coefficients for pigmented samples
pigmented <- colMeans(mod_mat[dds$colour %in% c("pink", "yellow"),])

# Our contrast of interest is
pigmented - white
```

Notice the contrast vector in this case assigns a "weight" of 0.5 to each of `colourpink` and `colouryellow`. This is equivalent to saying that we want to consider the average of pink and yellow expression. In fact, we could have also defined our contrast vector like this:

```{r}
# average of pink and yellow minus white
(pink + yellow)/2 - white
```

To obtain our results, we use the `results()` function as before:

```{r}
# get the results between pigmented and white
res2_pigmented <- results(dds, contrast = pigmented - white)
```


## Extra: why not define a new group in our design matrix?

For this last example (pigmented vs white), we may have considered creating a new variable in our column data:

```{r}
dds$pigmented <- factor(dds$colour %in% c("pink", "yellow"))
colData(dds)
```

and then re-run DESeq with a new design:

```{r, message=FALSE}
design(dds) <- ~ 1 + pigmented
dds <- DESeq(dds)
resultsNames(dds)
res1_pigmented <- results(dds, contrast = list("pigmented_TRUE_vs_FALSE"))
```

However, in this model the gene dispersion is estimated together for pink and yellow samples as if they were replicates of each other, which may result in inflated/deflated estimates. 
Instead, our approach above estimates the error within each of those groups. 

To check the difference one could compare the two approaches visually:

```{r, eval=FALSE}
# compare the log-fold-changes between the two approaches
plot(res1_pigmented$log2FoldChange, res2_pigmented$log2FoldChange)
abline(0, 1, col = "brown", lwd = 2)

# compare the errors between the two approaches
plot(res1_pigmented$lfcSE, res2_pigmented$lfcSE)
abline(0, 1, col = "brown", lwd = 2)
```


# Two factors with interaction (slide 7)

```{r}
# simulate data
dds <- makeExampleDESeqDataSet(n = 1000, m = 12, betaSD = 2)
dds$colour <- factor(rep(c("pink", "white"), each = 6))
dds$colour <- relevel(dds$colour, "white")
dds$condition <- factor(rep(c("sun", "shade"), 6))
dds <- dds[, order(dds$colour, dds$condition)]
colnames(dds) <- paste0("sample", 1:ncol(dds))
```

First let's look at our sample information:

```{r}
colData(dds)
```

This time we have two factors of interest, and we want to model both with an interaction (i.e. we assume that white and pink samples may respond differently to sun/shade). 
We define our design accordingly and fit the model:

```{r, message=FALSE}
design(dds) <- ~ 1 + colour + condition + colour:condition
dds <- DESeq(dds)
resultsNames(dds)
```

Because we have two factors and an interaction, the number of comparisons we can do is larger. 
Using our three-step approach from the model matrix, we do things exactly as we've been doing so far:

```{r}
# get the model matrix
mod_mat <- model.matrix(design(dds), colData(dds))

# Define coefficient vectors for each condition
pink_shade <- colMeans(mod_mat[dds$colour == "pink" & dds$condition == "shade", ])
pink_sun <- colMeans(mod_mat[dds$colour == "pink" & dds$condition == "sun", ])
white_shade <- colMeans(mod_mat[dds$colour == "white" & dds$condition == "shade", ])
white_sun <- colMeans(mod_mat[dds$colour == "white" & dds$condition == "sun", ])
```

We are now ready to define any contrast of interest from these vectors (for completeness we show the equivalent syntax using the coefficient's names from DESeq).

Pink vs White (in the shade):

```{r}
res1 <- results(dds, contrast = pink_shade - white_shade)
# or equivalently
res2 <- results(dds, contrast = list("colour_pink_vs_white"))
```

Pink vs White (in the sun):

```{r}
res1 <- results(dds, contrast = pink_sun - white_sun)
# or equivalently
res2 <- results(dds, contrast = list(c("colour_pink_vs_white",
                                       "colourpink.conditionsun")))
```

Sun vs Shade (for whites):

```{r}
res1 <- results(dds, contrast = white_sun - white_shade)
# or equivalently
res2 <- results(dds, contrast = list(c("condition_sun_vs_shade")))
```

Sun vs Shade (for pinks):

```{r}
res1 <- results(dds, contrast = pink_sun - pink_shade)
# or equivalently
res2 <- results(dds, contrast = list(c("condition_sun_vs_shade", 
                                       "colourpink.conditionsun")))
```

Interaction between colour and condition (i.e. do pinks and whites respond differently to the sun?):

```{r}
res1 <- results(dds, 
                contrast = (pink_sun - pink_shade) - (white_sun - white_shade))
# or equivalently
res2 <- results(dds, contrast = list("colourpink.conditionsun"))
```

In conclusion, although we can define these contrasts using DESeq coefficient names, it is somewhat more explicit (and perhaps intuitive?) what it is we're comparing using matrix-based contrasts. 


# Three factors, with nesting (slide 8)

```{r}
# simulate data
dds <- makeExampleDESeqDataSet(n = 1000, m = 24, betaSD = 2)
dds$colour <- factor(rep(c("white", "pink"), each = 12))
dds$colour <- relevel(dds$colour, "white")
dds$species <- factor(rep(LETTERS[1:4], each = 6))
dds$condition <- factor(rep(c("sun", "shade"), 12))
dds <- dds[, order(dds$colour, dds$species, dds$condition)]
colnames(dds) <- paste0("sample", 1:ncol(dds))
```

First let's look at our sample information:

```{r}
colData(dds)
```

Now we have three factors, but species is _nested_ within colour (i.e. a species is either white or pink, it cannot be both). 
Therefore, colour is a linear combination with species (or, another way to think about it is that colour is redundant with species). 
Because of this, we will define our design without including "colour", although later we can compare groups of species of the same colour with each other.

```{r, message=FALSE}
design(dds) <- ~ 1 + species + condition + species:condition
dds <- DESeq(dds)
resultsNames(dds)
```

Now it's harder to define contrasts between groups of species of the same colour using DESeq's coefficient names (although still possible). 
But using the model matrix approach, we do it in exactly the same way we have done so far!

Again, let's define our groups from the model matrix:

```{r}
# get the model matrix
mod_mat <- model.matrix(design(dds), colData(dds))

# define coefficient vectors for each group
pink_shade <- colMeans(mod_mat[dds$colour == "pink" & dds$condition == "shade", ])
white_shade <- colMeans(mod_mat[dds$colour == "white" & dds$condition == "shade", ])
pink_sun <- colMeans(mod_mat[dds$colour == "pink" & dds$condition == "sun", ])
white_sun <- colMeans(mod_mat[dds$colour == "white" & dds$condition == "sun", ])
```

It's worth looking at some of these vectors, to see that they are composed of weighted coefficients from different species. 
For example, for "pink" species, we have equal contribution from "speciesC" and "speciesD":

```{r}
pink_shade
```

And so, when we define our contrasts, each species will be correctly weighted:

```{r}
pink_sun - pink_shade
```

We can set our contrasts in exactly the same way as we did in the previous section (for completeness, we also give the contrasts using DESeq's named coefficients). 

Pink vs White (in the shade):

```{r, eval=FALSE}
res1_pink_white_shade <- results(dds, contrast = pink_shade - white_shade)
# or equivalently
res2_pink_white_shade <- results(dds, 
                                 contrast = list(c("species_B_vs_A"),
                                                 c("species_C_vs_A",
                                                   "species_D_vs_A")))
```

Pink vs White (in the sun):

```{r}
res1_pink_white_sun <- results(dds, contrast = pink_sun - white_sun)
# or equivalently
res2_pink_white_sun <- results(dds, 
                           contrast = list(c("species_B_vs_A",
                                             "speciesB.conditionsun"),
                                           c("species_C_vs_A", 
                                             "species_D_vs_A",
                                             "speciesC.conditionsun",
                                             "speciesD.conditionsun")))
```

And so on, for other contrasts of interest...


## Extra: imbalanced design

Let's take our previous example, but drop one of the samples from the data, so that we only have 2 replicates for it.

```{r, message=FALSE}
dds <- dds[, -13] # drop one of the species C samples
dds <- DESeq(dds)
resultsNames(dds)
```

Define our model matrix and coefficient vectors:

```{r}
mod_mat <- model.matrix(design(dds), colData(dds))
mod_mat

# define coefficient vectors for each group
pink_shade <- colMeans(mod_mat[dds$colour == "pink" & dds$condition == "shade", ])
white_shade <- colMeans(mod_mat[dds$colour == "white" & dds$condition == "shade", ])
pink_sun <- colMeans(mod_mat[dds$colour == "pink" & dds$condition == "sun", ])
white_sun <- colMeans(mod_mat[dds$colour == "white" & dds$condition == "sun", ])
```

Now let's check what happens to the pink_shade group:

```{r}
pink_shade
```

Notice that whereas before "speciesC" and "speciesD" had each a weight of 0.5, now they have different weights. 
That's because for speciesC there's only 2 replicates.
So, we have a total of 5 white individuals in the shade (2 from species C and 3 from D). 
Therefore, when we calculate the average coefficients for pinks, we need to do it as 0.4 x speciesC + 0.6 x speciesD. 

The nice thing about this approach is that we do not need to worry about any of this, the weights come from our `colMeans()` call automatically. 
And now, any contrasts that we make will take these weights into account:

```{r}
# pink vs white (in the shade)
pink_shade - white_shade

# interaction
(pink_sun - pink_shade) - (white_sun - white_shade)
```


# Further reading

- Forum discussion about nested design: http://seqanswers.com/forums/showthread.php?t=47766

